Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-03-06 numerics_paper/version2/numerical_chj.tex
last revision

\newcommand{\out}{\textup{out}}
\newcommand{\jet}{{\textup{jet}}}
\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\Ft}{\tilde{F}}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{f}}
\newcommand{\bfr}{{r}}
\newcommand{\bfF}{{F}}
\newcommand{\bfu}{{u}}
\newcommand{\entvar}{\eta'}
\newcommand{\bfq}{{Q}}
\newcommand{\bfx}{{x}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}
\newcommand{\efp}{\psi}
\newcommand{\entflux}{g}
\newcommand{\Fr}{{\rm Fr}}

@@

\begin{abstract}
We investigate the numerical artifact known as a carbuncle, in the solution
of the shallow water equations.  We propose a new Riemann solver that is based
on a local measure of the entropy residual and aims to avoid carbuncles while
maintaining high accuracy.
We propose a new challenging test problem for shallow water codes, 
consisting of a steady circular hydraulic
jump that can be physically unstable.  We show that numerical methods are prone to
either suppress the instability completely or form carbuncles.
We test existing cures for the carbuncle and show that only the proposed
method is able to avoid unphysical carbuncles without suppressing the
physical instability.
\end{abstract}

@@

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and LLF suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical}.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.

The most common test problems used to investigate carbuncle formation are
that of bow shock formation or a steady, grid-aligned planar shock.
In both of these problems, the correct behavior is the formation of a stable
shock profile without carbuncles.  This is achieved by certain methods
designed specifically to avoid carbuncles, but also by by typical first-order
accurate methods.  Thus these problems are not adequate on their own to
evaluate methods for practical calculations.
Elling \cite{elling2009carbuncle} proposed instead a problem specifically
designed to feature a carbuncle as the physical solution.  This has
been used as a test problem to identify methods that impose excessive dissipation.

@@

In this work, we describe and study the circular hydraulic jump as a new
test problem for shallow water discretizations.

\subsection{Outline}
In Section \ref{sec2}, we review some existing numerical methods for
the shallow water equations, focusing on certain Riemann solvers.
In Section \ref{sec:blended_rs}, we propose a new Riemann solver that blends
those of Roe and Rusanov in order to avoid carbuncles with being
excessively dissipative.  In Section \ref{sec:num}, we use Clawpack to
compare the performance of the new Riemann solver to existing
solvers on several standard shallow water test problems.  In Section \ref{sec:chj}, we
study the circular hydraulic jump using the newly proposed solver.
We find that although some
existing methods behave acceptably on previous test problems, they
are not capable of providing accurate solutions for the circular hydraulic jump
across the range of flow regimes we study.
Some conclusions and future directions are discussed in Section \ref{sec:conclusion}.

\section{Numerical methods for the shallow water equations} \label{sec2}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + \nabla \cdot f(q)_x & = 0,
\end{align}
where $q=[h, hu, hv]^T$ and the flux functions $f$ is defined in accordance
with \eqref{eq:sw}.

@@

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
Rusanov's solver approximates the Riemann solution by with a single wave traveling in each
direction.  Both waves are assumed to travel with speed $\lambda^{\max}_\imh$, which is an
upper bound on the wave speeds appearing in the true solution of the Riemann problem.
In this work, we take $\lambda_{i-1/2}^{\max}$ to be the upper bound
from \cite[Prop. 3.7]{azerad2017well}.
The waves are then given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfq_{i-1/2}-\bfq_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfq_{i}-\bfq_{i-1/2},
\end{align}
where the intermediate state $\bfu_{i-1/2}$ is determined by imposing conservation:
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfq_{i})-\bff(\bfq_{i-1}).
\end{align}
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^{p,\Roe}=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \sum_p\W_{i-1/2}^{p,\Roe} = \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfq_i-\bfq_{i-1}.
\end{align}

\subsection{Kemm's Riemann solver} \label{sec:kemm}
Several specialized Riemann solvers have been proposed in order to deal
with carbuncles.  Among them, only that proposed by Kemm has been implemented
and tested on the shallow water equations \cite{kemm2014note,bader2014carbuncle}.
The idea behind that solver, known as the HLLEMCC solver, is to combine
the HLLE and Roe solvers in such a way that the resulting method behaves
like Roe's when away from potential carbuncle instabilities, and like HLLE
where the potential for such instabilities arises.   This is done using
an indicator function based on the local residual of the Rankine-Hugoniot
condition for the shear wave.  The method requires some parameters that
may need to be tuned for a specific problem.  We will consider this method
in the numerical tests below.

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\subsection{The correction fluxes}
In this section we explain the choice of wave speeds \eqref{s_p}.
For the moment, we consider \eqref{blended-flux-1}-\eqref{lambda_p} without
entropy stabilization; i.e. we set $\lambda^{\min}_\iph=0$
for now.
We use $\W^{p,\Rus}$ to denote the waves in the Rusanov
solver and $\W^{p,\Roe}$ to denote the waves in the Roe solver.
The first-order method given by \eqref{first-order_via_fluct} with the numerical
flux \eqref{blended-flux-1} can also be written in
terms of fluctuations:
\begin{subequations}\label{ev_fluctuations}
\begin{align}
  \apdq_\imh & = \frac{1}{2}\sum_p \left(\hat\lambda_{i-1/2}^p + \lambda_{i-1/2}^p\right)\W_{i-1/2}^{p,\Roe}, \\
  \amdq_\iph & = \frac{1}{2}\sum_p \left(\hat\lambda_{i+1/2}^p - \lambda_{i+1/2}^p\right)\W_{i+1/2}^{p,\Roe}.
\end{align}
\end{subequations}
To implement the correction fluxes required for the second-order scheme \eqref{second-order_via_fluct},
we must also define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations},
and we instead use \eqref{s_p}.
Using this in \eqref{correction-fluxes} and \eqref{ev_fluctuations} in 
\eqref{second-order_via_fluct}, we obtain (in the absence of limiting)
the second-order scheme
\begin{align}\label{LWL}
  \bfq_i^{n+1}=\bfq_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^{p,\Roe}
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^{p,\Roe}\right],
\end{align}
It is clear again that if $\theta_{i\pm 1/2}=0$ we recover the standard
second-order Lax-Wendroff method based on Roe's Riemann solver.
We now show that if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method
based on Rusanov's Riemann solver.

>>>>

\newcommand{\out}{\textup{out}}
\newcommand{\jet}{{\textup{jet}}}
\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\Ft}{\tilde{F}}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{f}}
\newcommand{\bfr}{{r}}
\newcommand{\bfF}{{F}}
\newcommand{\bfu}{{u}}
\newcommand{\entvar}{\eta'}
\newcommand{\bfq}{{Q}}
\newcommand{\bfx}{{x}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}
\newcommand{\EV}{{\rm EV}}
\newcommand{\efp}{\psi}
\newcommand{\entflux}{g}
\newcommand{\Fr}{{\rm Fr}}

@@

\begin{abstract}
We investigate the numerical artifact known as a carbuncle, in the solution
of the shallow water equations.  We propose a new Riemann solver that is based
on a local measure of the entropy residual and aims to avoid carbuncles while
maintaining high accuracy.
We propose a new challenging test problem for shallow water codes, 
consisting of a steady circular hydraulic
jump that can be physically unstable.  We show that numerical methods are prone to
either suppress the instability completely or form carbuncles.
We test existing cures for the carbuncle. In our experiments, only the proposed
method is able to avoid unphysical carbuncles without suppressing the
physical instability.
\end{abstract}

@@

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and Rusanov suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical}.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solvers are used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe
based on a local measure of the entropy residual, 
as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.

The most common test problems used to investigate carbuncle formation are
that of bow shock formation or a steady, grid-aligned planar shock.
In both of these problems, the correct behavior is the formation of a stable
shock profile without carbuncles.  This is achieved by certain methods
designed specifically to avoid carbuncles, but also by typical first-order
accurate methods.  Thus these test problems are not adequate on their own to
evaluate methods for practical calculations.
Elling \cite{elling2009carbuncle} proposed instead a problem specifically
designed to feature a carbuncle as the physical solution.  This has
been used as a test problem to identify methods that impose excessive dissipation.

@@

In this work, we describe and study the circular hydraulic jump as a new
test problem for shallow water discretizations.
We solve this test problem using a novel Riemann solver that supresses the 
formation of carbuncles without dissipating important features of the solution.

\subsection{Outline}
In Section \ref{sec2}, we review some existing numerical methods for
the shallow water equations, focusing on certain Riemann solvers.
In Section \ref{sec:blended_rs}, we propose a new Riemann solver that blends
those of Roe and Rusanov in order to avoid carbuncles without being
excessively dissipative.  In Section \ref{sec:num}, we use Clawpack to
compare the performance of the new Riemann solver to existing
solvers on several standard shallow water test problems.  In Section \ref{sec:chj}, we
study the circular hydraulic jump using the newly proposed solver.
We find that although some
existing methods behave acceptably on previous test problems, they
are not capable of providing accurate solutions for the circular hydraulic jump
across the range of flow regimes we study.
Some conclusions and future directions are discussed in Section \ref{sec:conclusion}.

\section{Numerical methods for the shallow water equations} \label{sec2}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + \nabla \cdot f(q) & = 0,
\end{align}
where $q=[h, hu, hv]^T$ and the flux functions $f$ is defined in accordance
with \eqref{eq:sw}.

@@

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but 
(due to the use of excessive numerical dissipation)
is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
Rusanov's solver approximates the Riemann solution with a single wave traveling in each
direction.  Both waves are assumed to travel with speed $\lambda^{\max}_\imh$, which is an
upper bound on the wave speeds appearing in the true solution of the Riemann problem.
In this work, we take $\lambda_{i-1/2}^{\max}$ to be the upper bound
from \cite[Prop. 3.7]{azerad2017well}.
The waves are then given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfq_{i-1/2}-\bfq_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfq_{i}-\bfq_{i-1/2},
\end{align}
where the intermediate state $\bfu_{i-1/2}$ is determined by imposing conservation:
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfq_{i})-\bff(\bfq_{i-1}).
\end{align}
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^{p,\Roe}=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \sum_p\W_{i-1/2}^{p,\Roe} = \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfq_i-\bfq_{i-1}.
\end{align}

\subsection{Kemm's Riemann solver} \label{sec:kemm}
Several specialized Riemann solvers have been proposed in order to deal
with carbuncles.  Among them, only that proposed by Kemm has been implemented
and tested on the shallow water equations \cite{kemm2014note,bader2014carbuncle}.
The idea behind that solver, known as the HLLEMCC solver, is to combine
the HLLE and Roe solvers in such a way that the resulting method behaves
like Roe's away from potential carbuncle instabilities, and like HLLE
where the potential for such instabilities arises.   This is done using
an indicator function based on the local residual of the Rankine-Hugoniot
condition for the shear wave.  The method requires some parameters that
may need to be tuned for a specific problem.  We will consider this method
in the numerical tests below.

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Recall that $Q=[h,hu,hv]^t$. We use the total energy as entropy; that is, we consider 
\begin{align*}
  \eta(\bfq)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(\bfq)=
    \frac{\eta(\bfq)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\subsection{The correction fluxes}
In this section we explain the choice of wave speeds \eqref{s_p}.
For the moment, we consider \eqref{blended-flux-1}-\eqref{lambda_p} without
entropy stabilization; i.e., we set $\lambda^{\min}_\iph=0$
for now.
We use $\W^{p,\Rus}$ to denote the waves in the Rusanov
solver and $\W^{p,\Roe}$ to denote the waves in the Roe solver.
The first-order method given by \eqref{first-order_via_fluct} with the numerical
flux \eqref{blended-flux-1} can also be written in
terms of fluctuations:
\begin{subequations}\label{ev_fluctuations}
\begin{align}
  \apdq_\imh & = \frac{1}{2}\sum_p \left(\hat\lambda_{i-1/2}^p + \lambda_{i-1/2}^p\right)\W_{i-1/2}^{p,\Roe}, \\
  \amdq_\iph & = \frac{1}{2}\sum_p \left(\hat\lambda_{i+1/2}^p - \lambda_{i+1/2}^p\right)\W_{i+1/2}^{p,\Roe}.
\end{align}
\end{subequations}
To implement the correction fluxes required for the second-order scheme \eqref{second-order_via_fluct},
we must also define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations},
and we instead use \eqref{s_p}.
Using this in \eqref{correction-fluxes} and \eqref{ev_fluctuations} in 
\eqref{second-order_via_fluct}, we obtain (in the absence of limiting)
the second-order scheme
\begin{align}\label{LWL}
  \bfq_i^{n+1}=\bfq_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^{p,\Roe}
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^{p,\Roe}\right],
\end{align}
It is clear again that if $\theta_{i\pm 1/2}=0$ we recover the standard
second-order Lax-Wendroff method based on Roe's Riemann solver.
We now show that if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method
based on Rusanov's Riemann solver.
